---
title: "Time Series Project"
subtitle: "Spring 2024"
author: "Nicholas Sager, Anishka Peter"
date: last-modified
format: 
  html: 
    toc: true
    toc-location: left
    toc-depth: 4
    embed-resources: true
    # self-contained-math: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(tswge)
library(nnfor)
library(vars)
# library(tseries)
# library(orcutt)
```

## Data
Introduce dataset, response variable, and scenario.
ACFs and Spectral Density plots.
```{r}

```


## Models
### Univariate: At least 2 candidate ARMA / ARIMA models
a.	The models in factored form or at least separate the stationary and non-
stationary factors with standard deviation or variance of the white noise.
b.	AIC
c.	ASE (short and long term forecasts)
d.	Rolling Window RMSE (short and long term forecasts)
e.	At least 10 superimposed spectral densities from 10 generatied 
realizations like we did in Unit 11.  Use these to help choose between the at least two candidate univariate models. 
f.	Visualization of Forecasts for both the short- and long-term Horizons. 
g.	Be sure and include confidence intervals when possible (I don’t have 
code for confidence intervals from MLP models at the moment… but that would be a good thing to work on!  )

### Multivariate: At least one multivariate model (VAR or MLR with Correlated Errors)
i.	Include an ASE (rolling window is not yet available in multivariate models)
  1.	Short Horizon (you pick the length.. could be one step ahead)
  2.	Long Horizon (you pick ... just must be longer than the short horizon.)
ii.	Describe the explanatory variable(s) used in the model and why you felt they were significant / important.
iii.	Visualization of Forecasts for both the short- and long-Horizons.
iv.	Be sure and include confidence intervals if using VAR … 

### MLP Model
i.	ASE (short and long term forecasts)
ii.	Rolling Window RMSE (short and long term forecasts (only if univariate)
iii.	Visualization of Forecasts for both the short- and long-term Horizons. 
iv.	Confidence / Prediction intervals are not required (I don’t have 
code for confidence / prediction intervals (bootstrap intervals) for MLP models at the moment… but that would be a good thing to work on!  )

### Ensemble Model

## Model Comparison and Final Forecasts
i.	Provide a table comparing all models on at least ASE and rwRMSE (if available).
ii.	Include at least one ensemble model in addition to the models above.
iii.	Make a case as to why you feel one of your candidate models is the most useful.
iv.	Provide you final short and long term forecasts with that model. 


## Conclusion
